### 2021-01-21 ch02_scientific_background/scientific_background.tex
Added additional section to chapter 04

This chapter introduces the scientific background of this thesis. First, basic
concepts of climate model simulations and associated uncertainties are
introduced. Next, important metrics describing climate sensitivity and
fundamental biogeochemical processes of the global carbon cycle are presented.
Finally, state-of-the art techniques that can be used to reduce uncertainties
in projections of the future climate with observations are shown. These methods
form the basis for new techniques developed in this thesis.

>>>>

This chapter introduces the scientific background of this thesis. First, basic
concepts of climate model simulations and associated uncertainties are
introduced (\cref{sec:02:esms}). Next, important metrics describing climate
sensitivity (\cref{sec:02:climate_sensitivity}) and fundamental biogeochemical
processes of the global carbon cycle (\cref{sec:02:esms}) are presented.
Finally, state-of-the art techniques that can be used to reduce uncertainties
in projections of the future climate with observations are shown
(\cref{sec:02:reducing_uncertainties}). These methods form the basis for new
techniques developed in this thesis.

### 2021-01-21 ch03_esmvaltool/esmvaltool.tex
Added additional section to chapter 04

In order to answer the research questions of this thesis posed in
\cref{sec:01:research_questions}, a reliable and efficient tool to read,
process and evaluate climate model output and observational data is necessary.
A valuable software that fits these criteria is the \ac{ESMValTool}. The
\ac{ESMValTool} is an open-source community diagnostics and performance metrics
tool for the routine evaluation of \ac{ESM} output, which notably facilitates
the analysis of \ac{CMIP} models (\url{https://www.esmvaltool.org}). For this
reason, all analyses presented in
\cref{ch:04:papers_ecs_tcr_assessment,ch:05:paper_ecs,ch:06:paper_gpp} of this
thesis were implemented into the \ac{ESMValTool}. Apart from that, further
substantial changes and additions to the code base of the \ac{ESMValTool} have
been contributed that improve the routine evaluation of climate models, which
is beneficial for the entire scientific community. This lead to co-authorship
in the scientific documentation of the \ac{ESMValTool}, which is published in
four peer-reviewed studies \autocite{Eyring2020, Lauer2020, Righi2020,
  Weigel2020}. After a brief overview over the \ac{ESMValTool}'s structure,
this chapter presents all these contributions that are not documented in other
chapters of this thesis.

>>>>

In order to answer the research questions of this thesis posed in
\cref{sec:01:research_questions}, a reliable and efficient tool to read,
process and evaluate climate model output and observational data is necessary.
A valuable software that fits these criteria is the \ac{ESMValTool}. The
\ac{ESMValTool} is an open-source community diagnostics and performance metrics
tool for the routine evaluation of \ac{ESM} output, which notably facilitates
the analysis of \ac{CMIP} models (\url{https://www.esmvaltool.org}). For this
reason, all analyses presented in
\cref{ch:04:papers_ecs_tcr_assessment,ch:05:paper_ecs,ch:06:paper_gpp} of this
thesis were implemented into the \ac{ESMValTool}. Apart from that, further
substantial changes and additions to the code base of the \ac{ESMValTool} have
been contributed that improve the routine evaluation of climate models, which
is beneficial for the entire scientific community. This lead to co-authorship
in the scientific documentation of the \ac{ESMValTool}, which is published in
four peer-reviewed studies \autocite{Eyring2020, Lauer2020, Righi2020,
  Weigel2020}. After a brief overview of the \ac{ESMValTool}'s structure
(\cref{sec:03:esmvaltool}), this chapter presents all these contributions that
are not documented in other chapters of this thesis
(\cref{sec:03:contributions_to_esmvalcore,sec:03:contributions_to_esmvaltool}).

### 2021-01-21 ch04_papers_ecs_tcr_assessment/papers_ecs_tcr_assessment.tex
Added additional section to chapter 04

In order to reduce uncertainties in multi-model climate projections, a first
important step is the assessment of the desired target variables in the
corresponding climate model ensemble. This is particularly relevant when a new
generation of \acp{ESM} is published that includes considerable modifications
compared to the respective predecessor model versions. Currently, we are in a
situation like this with the \acs{CMIP}6 ensemble, for which new data is still
released every day (as of \TheMonth{}). In light of the upcoming international
climate assessment of the \ac{IPCC}'s \acs{AR}6, the evaluation of
policy-relevant metrics like \ac{ECS} or \ac{TCR} for these new \acs{CMIP}6
models is crucial since they provide vital information about the future climate
of planet Earth. As the \ac{ESMValTool} is an excellent tool that allows for a
quick and robust evaluation of new \ac{CMIP} data as soon as it gets published
on the \ac{ESGF} servers (see \cref{sec:03:esmvaltool}), we use it to assess
and analyze climate sensitivity (expressed by \ac{ECS} and \ac{TCR}) of the
\acs{CMIP}6 models and compare it to corresponding results of predecessor model
generations. This work is already published in two scientific papers:
\textcite{Bock2020} and \textcite{Meehl2020}. For \textcite{Bock2020}, the
author of this thesis contributed their figure 8 (a bar chart showing the
\ac{ECS} for the three climate model generations \acs{CMIP}3, \acs{CMIP}5 and
\acs{CMIP}6), their figure 10 (map plots showing cloud-related feedback
parameters for the \acs{CMIP}5 and \acs{CMIP}6 \acp{MMM}; see
\cref{fig:04:cloud_feedback_parameters}) and discussion about \ac{ECS} of the
\acs{CMIP}6 models. For \textcite{Meehl2020}, the author of this thesis
contributed all figures and tables of the study and details on the calculations
of \ac{ECS} and \ac{TCR}.

\section{Evaluation of \acs{ECS} and \acs{TCR} in \acs{CMIP}6}
\label{sec:04:evaluation_ecs_and_tcr}

Following \cref{subsec:02:ecs}, \ac{ECS} is calculated with the Gregory
regression method using \ac{GSAT} and \ac{TOA} net radiation data for 150 years
of a \nxcotwo{4} simulation \autocite{Gregory2004}. This calculation is
illustrated in \cref{fig:04:gregory_regression_cmip6_mmm} for the \acs{CMIP}6
\ac{MMM}, which yields an \ac{ECS} of $3.74 \unit{K}$ when all 150 years of the
run are used. However, similar to the \acs{CMIP}5 ensemble, the exact value of
\ac{ECS} depends on the years considered in the Gregory regression (see
\cref{fig:02:gregory_regression_different_years}). Using only the first 20
years of the simulation gives a significantly lower \ac{ECS} of $3.31 \unit{K}$
than using only the last 130 years of the simulation, which gives an \ac{ECS}
of $4.05 \unit{K}$. As thoroughly describes in \cref{subsec:02:ecs}, the reason
for this is the state and time dependence of the climate feedback parameter,
which is given by the slope of the Gregory regression line. Due to non-linear
effects in the feedbacks, this slope changes over time, resulting in lower
(higher) values of \ac{ECS} when only early (late) years of the simulation are
considered.

@@

In contrast to that, the transient response of the climate system \ac{TCR} is
calculated from \ac{GSAT} data at the time of \ac{CO2} doubling in a
\onepctcotwo{} run \commentcite{Bindoff2013}{see \cref{subsec:02:tcr}}. Using
these two definitions and the \ac{ESMValTool}, we assess \ac{ECS} and \ac{TCR}
for all \acs{CMIP}5 and \acs{CMIP}6 models where the necessary temperature and
radiation data is available (as of \TheMonth{}).
\Cref{tab:04:ecs_tcr_cmip5,tab:04:ecs_tcr_cmip6} give an overview over the
results for the \acs{CMIP}5 and \acs{CMIP}6 models, respectively.

@@

@@

To illustrate this and put it into historical context,
\cref{fig:04:historical_ecs_tcr_ranges} shows the assessed \ac{ECS} and
\ac{TCR} ranges over the years from the Charney report \autocite{Charney1979}
and the different \acp{AR} of the \ac{IPCC} in combination with the
corresponding modeled ranges from the different \ac{CMIP} generations. Since
the Charney report in 1979, the assessed range of \ac{ECS} of $\range{1.5}{4.5}
\unit{K}$ has almost remained unchanged during the last 40 years
\autocite{Charney1979, Mitchell1990, Kattenberg1996, Albritton2001,
  Stocker2013} with the exception of \acs{AR}4, where the lower bound was
temporarily increased to $2.0 \unit{K}$ \autocite{Solomon2007}. For \acs{AR}2,
\acs{AR}4 and \acs{AR}5, the corresponding climate model generations from
\acs{CMIP}1, \acs{CMIP}3 and \acs{CMIP}5 more or less agree with this assessed
range \autocite{Kattenberg1996, Randall2007, Flato2013}. In contrast to that,
some of the early climate models used in \acs{AR}1 and some \acs{CMIP}2 models
used in \acs{AR}3 exhibit values of \acs{ECS} well above $4.5 \unit{K}$,
resulting in a upper model range of about $0.6 \unit{K}$ higher than the
assessed range \autocite{Mitchell1990, Cubasch2001}. However, these deviations
from the assessed range are insignificant to the ones present in the
\acs{CMIP}6 ensemble, which have become evident over the course of 2020. As
shown in \cref{fig:04:historical_ecs_tcr_ranges}, both the lower and upper
bound are more extreme than in any previous climate model generation, resulting
in a \acs{CMIP}6  \ac{ECS} model range of $\range{1.8}{5.6} \unit{K}$ (as of
\TheMonth{}). This increase in range is particularly relevant for the upper
bound, which is about $1.1 \unit{K}$ higher than the assessed upper bound of
$4.5 \unit{K}$. Moreover, it is not only a small fraction of models that
exceeds this upper bound: out of the 42 analyzed \ac{CMIP}6 models, a third (14
models) has an \ac{ECS} above $4.5 \unit{K}$. In addition to our analysis,
several modeling centers independently confirmed high \ac{ECS} values in their
models \autocite{Andrews2019, Gettelman2019, Wyser2020}. A list of possible
reasons for the increased climate sensitivity in the \acs{CMIP}6 ensemble
including aerosol-cloud interactions and changes in the shortwave cloud
feedback over the Southern Ocean is given in
\cref{sec:04:possible_reasons_high_ecs_cmip6}.

@@

\begin{figure}[!t]
  \centering
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/abs_cmip6_netcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:a}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/bias_cmip6_netcre-cmip5_netcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:b}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/abs_cmip6_swcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:c}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/bias_cmip6_swcre-cmip5_swcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:d}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/abs_cmip6_lwcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:e}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/bias_cmip6_lwcre-cmip5_lwcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:f}
  \end{subfigure}
  \caption{Geographical distributions of the net (a), shortwave (c) and
    longwave (e) \acf{CRE} feedback parameters for the \acs{CMIP}6 \acf{MMM}
    and the corresponding differences to the \acs{CMIP}5 \acs{MMM}, again for
    the net (b), shortwave (d) and longwave (f) components. The \acs{CRE}
    feedback parameters for each grid cell are calculated with a Gregory
    regression of the grid cell's \acs{CRE} versus the change in the
    \acl{GSAT} (see \cref{subsec:02:cloud_feedback_parameters}). Updated from
    \textcite{Bock2020}.}
  \label{fig:04:cloud_feedback_parameters}
\end{figure}

>>>>

In order to reduce uncertainties in multi-model climate projections, a first
important step is the assessment of the desired target variables in the
corresponding climate model ensemble. This is particularly relevant when a new
generation of \acp{ESM} is published that includes considerable modifications
compared to the respective predecessor model versions. Currently, we are in a
situation like this with the \acs{CMIP}6 ensemble, for which new data is still
released every day (as of \TheMonth{}). In light of the upcoming international
climate assessment of the \ac{IPCC}'s \acs{AR}6, the evaluation of
policy-relevant metrics like \ac{ECS} or \ac{TCR} for these new \acs{CMIP}6
models is crucial since they provide vital information about the future climate
of planet Earth. As the \ac{ESMValTool} is an excellent tool that allows for a
quick and robust evaluation of new \ac{CMIP} data as soon as it gets published
on the \ac{ESGF} servers (see \cref{sec:03:esmvaltool}), we use it to assess
and analyze \ac{ECS} and \ac{TCR} of the \acs{CMIP}6 models and compare it to
corresponding results of predecessor model generations and international
climate assessments. This work is already published in two scientific papers:
\textcite{Bock2020} and \textcite{Meehl2020}. For \textcite{Bock2020}, the
author of this thesis contributed their figure 8 (a bar chart showing the
\ac{ECS} for the three climate model generations \acs{CMIP}3, \acs{CMIP}5 and
\acs{CMIP}6), their figure 10 (map plots showing cloud-related feedback
parameters for the \acs{CMIP}5 and \acs{CMIP}6 \acp{MMM}; see
\cref{fig:04:cloud_feedback_parameters}) and discussion about \ac{ECS} of the
\acs{CMIP}6 models. For \textcite{Meehl2020}, the author of this thesis
contributed all figures and tables of the study. This chapter first presents
the evaluation of climate sensitivity for the \acs{CMIP}5 and \acs{CMIP}6
ensembles (\cref{sec:04:evaluation_ecs_and_tcr}) and puts the corresponding
\ac{ECS} and \ac{TCR} values into historical context
(\cref{sec:04:historical_context}). Finally, possible reasons for the apparent
increase of climate sensitivity in the \acs{CMIP}6 models are briefly discussed
(\cref{sec:04:possible_reasons_high_ecs_cmip6}).

\section{Evaluation of \acs{ECS} and \acs{TCR} in \acs{CMIP}5 and \acs{CMIP}6}
\label{sec:04:evaluation_ecs_and_tcr}

Following \cref{subsec:02:ecs}, \ac{ECS} is calculated with the Gregory
regression method using \ac{GSAT} and \ac{TOA} net radiation data for 150 years
of a \nxcotwo{4} simulation \autocite{Gregory2004}. This calculation is
illustrated in \cref{fig:04:gregory_regression_cmip6_mmm} for the \acs{CMIP}6
\ac{MMM}, which yields an \ac{ECS} of $3.74 \unit{K}$ when all 150 years of the
run are used. However, similar to the \acs{CMIP}5 ensemble, the exact value of
\ac{ECS} depends on the years considered in the Gregory regression (see
\cref{fig:02:gregory_regression_different_years}). Using only the first 20
years of the simulation gives a significantly lower \ac{ECS} of $3.31 \unit{K}$
than using only the last 130 years of the simulation, which gives an \ac{ECS}
of $4.05 \unit{K}$. As thoroughly describes in \cref{subsec:02:ecs}, the reason
for this is the state and time dependence of the climate feedback parameter,
which is given by the slope of the Gregory regression line. Due to non-linear
effects in the feedbacks, this slope changes over time, resulting in lower
(higher) values of \ac{ECS} when only early (late) years of the simulation are
considered. In contrast to that, the transient response of the climate system
\ac{TCR} is calculated from \ac{GSAT} data at the time of \ac{CO2} doubling in
a \onepctcotwo{} run \commentcite{Bindoff2013}{see \cref{subsec:02:tcr}}. Using
these two definitions and the \ac{ESMValTool}, we assess \ac{ECS} and \ac{TCR}
for all \acs{CMIP}5 and \acs{CMIP}6 models where the necessary temperature and
radiation data is available (as of \TheMonth{}).
\Cref{tab:04:ecs_tcr_cmip5,tab:04:ecs_tcr_cmip6} give an overview over the
results for the \acs{CMIP}5 and \acs{CMIP}6 models, respectively.

@@

@@

\section{Comparison to Previous \acs{CMIP} Generations and International
  Climate Assessments}
\label{sec:04:historical_context}

To illustrate the results of the previous section and put them into historical
context, \cref{fig:04:historical_ecs_tcr_ranges} shows the assessed \ac{ECS}
and \ac{TCR} ranges over the years from the Charney report
\autocite{Charney1979} and the different \acp{AR} of the \ac{IPCC} in
combination with the corresponding modeled ranges from the different \ac{CMIP}
generations. Since the Charney report in 1979, the assessed range of \ac{ECS}
of $\range{1.5}{4.5} \unit{K}$ has almost remained unchanged during the last 40
years \autocite{Charney1979, Mitchell1990, Kattenberg1996, Albritton2001,
  Stocker2013} with the exception of \acs{AR}4, where the lower bound was
temporarily increased to $2.0 \unit{K}$ \autocite{Solomon2007}. For \acs{AR}2,
\acs{AR}4 and \acs{AR}5, the corresponding climate model generations from
\acs{CMIP}1, \acs{CMIP}3 and \acs{CMIP}5 more or less agree with this assessed
range \autocite{Kattenberg1996, Randall2007, Flato2013}. In contrast to that,
some of the early climate models used in \acs{AR}1 and some \acs{CMIP}2 models
used in \acs{AR}3 exhibit values of \acs{ECS} well above $4.5 \unit{K}$,
resulting in a upper model range of about $0.6 \unit{K}$ higher than the
assessed range \autocite{Mitchell1990, Cubasch2001}. However, these deviations
from the assessed range are insignificant to the ones present in the
\acs{CMIP}6 ensemble, which have become evident over the course of 2020. As
shown in \cref{fig:04:historical_ecs_tcr_ranges}, both the lower and upper
bound are more extreme than in any previous climate model generation, resulting
in a \acs{CMIP}6  \ac{ECS} model range of $\range{1.8}{5.6} \unit{K}$ (as of
\TheMonth{}). This increase in range is particularly relevant for the upper
bound, which is about $1.1 \unit{K}$ higher than the assessed upper bound of
$4.5 \unit{K}$. Moreover, it is not only a small fraction of models that
exceeds this upper bound: out of the 42 analyzed \ac{CMIP}6 models, a third (14
models) has an \ac{ECS} above $4.5 \unit{K}$. In addition to our analysis,
several modeling centers independently confirmed high \ac{ECS} values in their
models \autocite{Andrews2019, Gettelman2019, Wyser2020}. A list of possible
reasons for the increased climate sensitivity in the \acs{CMIP}6 ensemble
including aerosol-cloud interactions and changes in the shortwave cloud
feedback over the Southern Ocean is given in
\cref{sec:04:possible_reasons_high_ecs_cmip6}.

@@

@@

\begin{figure}[p]
  \centering
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/abs_cmip6_netcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:a}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/bias_cmip6_netcre-cmip5_netcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:b}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/abs_cmip6_swcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:c}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/bias_cmip6_swcre-cmip5_swcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:d}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/abs_cmip6_lwcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:e}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SmallSubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch04_papers_ecs_tcr_assessment/figs/bias_cmip6_lwcre-cmip5_lwcre.pdf}
    \caption{}
    \label{fig:04:cloud_feedback_parameters:f}
  \end{subfigure}
  \caption{Geographical distributions of the net (a), shortwave (c) and
    longwave (e) \acf{CRE} feedback parameters for the \acs{CMIP}6 \acf{MMM}
    and the corresponding differences to the \acs{CMIP}5 \acs{MMM}, again for
    the net (b), shortwave (d) and longwave (f) components. The \acs{CRE}
    feedback parameters for each grid cell are calculated with a Gregory
    regression of the grid cell's \acs{CRE} versus the change in the
    \acl{GSAT} (see \cref{subsec:02:cloud_feedback_parameters}). Updated from
    \textcite{Bock2020}.}
  \label{fig:04:cloud_feedback_parameters}
\end{figure}
